---
title: 'MECH481A6: Engineering Data Analysis in R'
subtitle: 'Chapter 10 Homework: Measurement' 
author: 'Flynn Nyman'
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: pdf_document
---

```{r global-options, include=FALSE}
# set global options for figures, code, warnings, and messages
knitr::opts_chunk$set(fig.width=6, fig.height=4, fig.path="../figs/",
                      echo=FALSE, warning=FALSE, message=FALSE)
```

# Load packages

```{r load-packages, message=FALSE}
# load packages for current session
library(tidyverse) 
library(gridExtra) 
library(MASS)
library(lubridate)
library(dplyr)
```

# Chapter 10 Homework

This homework will give you practice at working with a measurement dataset: `airlift_mass_repeatability.csv`. This data set represents repeated measures of "blank" air sampling filters. 

A couple notes to consider when reporting answers in response to questions. The microbalance used to make these measurements reads out to the nearest microgram ($\mu g$), which is 0.000001 $g$ or 0.001 $mg$. Thus, be careful when reporting descriptive statistics so as not to overstate your **precision**. Use the `round()` function to avoid reporting more than 0.1 $\mu g$ of precision (or 0.0001 $mg$). Here is some example code that uses the `across()` function from `dplyr::` to round numeric output to just four digits (appropriate for $mg$ units in this exercise):

`dplyr::mutate(across(.cols = where(is.numeric), .fns = round, 3))`

\newpage

## Question 1
Import the `airlift_mass_repeatability.csv` file into a data frame called `blanks` and perform the following data wrangling in a single pipe:  

- retain only the first 3 columns of data;
- rename the columns with the names `date`, `id`, and `mass_mg`;
- convert the `date` column vector into a date class object using `lubridate::`
- convert the `id` variable to a class `factor` (this can be accomplished using `base::as.factor()` or `purrr::as_factor()`)
- create a new column vector named `mass_mg` by rescaling the `mass_g` data (i.e., convert $g$ to $mg$ by multiplying `mass_g` by 1000)

```{r import-clean-data}
airlift_data <- read.csv("./AIRLIFT_mass_repeatability.csv") %>%
  dplyr::select(Date, Filter.ID, Mass..g.) %>%
  rename("date" = "Date", "id" = "Filter.ID", "mass_g" = "Mass..g.") %>%
  mutate(date = dmy(date)) %>%
  mutate(id = as.factor(id)) %>%
  mutate(mass_mg = mass_g*1000)
```

## Question 2:  

  2a. Are there any NAs present in the data frame?  
There are no missing values in the data set.  
  2b. How many unique filter IDs are present in this data frame?  
There are 5 unique filter IDs.  
  2c. How many samples are present for each filter ID? Hint: look up the `dplyr::count()` function.  
There are 78 occurrences of IDs 41666, 41667, 41668, and 41671. There are 76 occurrences of ID 41669.  
  2d. Over how long of a period were these blank measurements made? Hint: this can be done in base R with a `max() - min()` or with `lubridate::interval() %>% as.duration()`.
There is a span of 35 days between the earliest and latest measurement.  
```{r unique-IDs}
NA_present <- any(is.na(airlift_data)) %>%
  print()
## There are no missing values in the data set.

unique_IDs <- unique(airlift_data$id) %>%
  print()
## There are 5 unique filter IDs in the data set.

airlift_data %>%
  count(id)
## There are 78 occurrences of IDs 41666, 41667, 41668, and 41671. There are 76 occurrences of ID 41669.

measurement_span <- as.numeric(difftime(max(airlift_data$date, na.rm = TRUE), min(airlift_data$date, na.rm = TRUE), units = "days")) %>%
  print()
## There is a span of 35 days between the earliest and latest measurement.
```

## Question 3

Group the `blanks` data frame by `id` and calculate mean, median, and standard deviations for each filter id.  Hint: use `group_by() %>% summarise()` to do this efficiently.

```{r blank-descriprives}
summary <- airlift_data %>%
  group_by(id) %>%
  summarise(mean(mass_mg), median(mass_mg), sd(mass_mg)) %>%  #XL: If you can name mean(mass_mg), etc. in summarise, then you don't need to rename them. For example, summarise(mean_mg = mean(mass_mg), median_me = median(mass_mg), sd_mg = sd(mass_mg))
  rename('mean' = 'mean(mass_mg)',                            #XL: Additionally, please pay attention to the  **precision**. Use the `round()` function to avoid reporting more than 0.1 $\mu g$ of precision (or 0.0001 $mg$). For example, round(mean(mass_mg), digits = 3). Same for Question 3. 
         'median' = 'median(mass_mg)',
         'sd' = 'sd(mass_mg)') %>%
  print()
```

## Question 4

Calculate the limit of detection (LOD) for this measurement method. Note: you will need to calculate standard deviations for each filter `id` (as done in question 3) and then estimate LOD from $LOD = 3\cdot \sigma_b$ where $\sigma_b$ is calculated for each filter `id`.

```{r Limit of Detection}
LOD <- summary %>%
  mutate(LOD = mean + 3*sd) %>%       #XL: LOD = 3*sd, no relationship with mean value.
  dplyr::select(!c(id, mean, median, sd)) %>%   
  print()
```

\newpage

# Appendix

```{r show_code, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}
```
