---
title: 'MECH476: Engineering Data Analysis in R'
subtitle: 'Chapter 7 Homework: Multivariate Exploratory Data Analysis' 
author: 'Flynn Nyman'
date: '`r format(Sys.time(), "%d %B, %Y")`'
output: pdf_document
---

```{r global-options, include=FALSE}
# set global options for figures, code, warnings, and messages
knitr::opts_chunk$set(fig.width=6, fig.height=4, fig.path="../figs/",
                      echo=FALSE, warning=FALSE, message=FALSE)
```

# Load packages

```{r load-packages, message=FALSE}
# load packages for current session
library(tidyverse)
```

# Chapter 7 Homework

In [Chapter 5](https://smogdr.github.io/edar_coursebook/eda1.html#cumulative-distribution-plot), we briefly explored data on the salaries of engineering
graduates from the National Science Foundation 2017 [National Survey of College Graduates](https://www.nsf.gov/statistics/srvygrads/) from a univariate
perspective. Now, let's explore the relationships between multiple variables. 

When a question asks you to make a plot, remember to set a theme, title, 
subtitle, labels, colors, etc. It is up to you how to personalize your plots, 
but put in some effort, and make the plotting approach consistent throughout
the document. For example, you could use the same theme for all plots. I also
like to use the subtitle as a place for the main summary for the viewer.

\newpage

## Question 1: Data wrangling

Within a pipeline, import the data from the .csv file, convert all column names
to lowercase text (either "manually" with `dplyr::rename()`, or use
[`clean_names()` from the `janitor` package](https://rdrr.io/cran/janitor/man/clean_names.html)), convert `gender`
from "numeric" to "factor", and drop any and all observations with `salary` 
recorded as 0. Assign this to a dataframe object with a meaningful name. 

```{r import-data}
# import and tidy salary data
salaries <- read_csv("./ME_salaries.csv") %>%
  rename(salary = SALARY,
         age = AGE,
         gender = GENDER) %>%
  mutate(gender = factor(gender)) %>%
  subset(salary != 0)
```

How many observations have a 0 (zero) value for salary? Note: The last question asked you to remove these observations from the resultant data frame.

```{r salary-zero}
# number of observations with salary as 0 
nrow(read_csv("./ME_salaries.csv")) - nrow(salaries)
```

What are the levels in `gender`? (Ignore the fact that the observations
refer to "biological sex", not "gender". *Gender* is now recognized as a fluid
term with more than two options; *biological sex* - what was assigned at birth - 
is binary term).

```{r gender-levels}
# number of factor levels
unique(salaries$gender)   #XL: Please remember the new function we learnt in CH7, level()
```

## Question 2: Univariate EDA

Using what you learned in Chapter 5, generate basic plots and/or descriptive 
statistics to explore `age`, `gender`, and `salary`. List whether each variable
is continuous or categorical, and explain how and why you adjusted your EDA
approach accordingly.

```{r univariate-eda, include=FALSE}
# univariate eda
ggplot(salaries, aes(x = salary)) +
  geom_histogram(bins = 50, color = "black") +
  labs(x = "Salary", y = "Count", title = "Histogram of salaries of ME grads") +
  theme_minimal()

ggplot(salaries, aes(x = gender)) +
  geom_bar(fill = "darkgreen") +
  labs(x = "Gender", y = "Count", title = "Number of Engineers by Gender")
  theme_minimal()

ggplot(salaries, aes(x = age)) +
  geom_histogram(bins = 50, fill = "darkgreen", color = "black") +
  labs(x = "Age", y = "Count", title = "Age of ME Engineers") +
  theme_minimal()
```
Age and salary are both continuous variables as there are many different observations in the data set and no limit for how many options they’ve  can take on. Gender on the other hand is a caregorical variable as plot 2 shows that it only takes on two options in this dataset.

\newpage

## Question 3: Multivariate histograms

Create a histogram of `salary`, faceted by `gender`. Add `bins = 50` and 
`color = "lightgrey"`.

```{r hist-salary}
# histogram of salaries split by gender
ggplot(salaries, aes(x = salary, fill = gender)) +
  geom_histogram(bins = 50, color = "lightgrey") +           #XL: Please try to use facet, e.g., facet_wrap(~gender), even though fill = gender has the same function here.
  labs(x = "Salary", y = "Count", title = "Histogram of salaries of ME grads by gender") +
  theme_minimal()
```

Create a histogram of `age`, faceted by `gender`. Add `bins = 50` and 
`color = "lightgrey"`.

```{r hist-age}
# histogram of ages split by gender
ggplot(salaries, aes(x = age, fill = gender)) +             #XL: Please try to use facet, e.g., facet_wrap(~gender), even though fill = gender has the same function here.
  geom_histogram(bins = 50, color = "lightgrey") +
  labs(x = "Age", y = "Count", title = "Histogram of age of ME grads by gender") +
  theme_minimal()
```

\newpage

## Question 4: Multivariate boxplots

Create a boxplot of `salary`, faceted by `gender`. Use `oulier.shope = 1` to 
better visualize the outliers.

```{r salary-box}
# boxplots of salary data by gender
ggplot(salaries, aes(y = salary, fill = gender)) +       #XL: Please try to use facet, e.g., facet_wrap(~gender), even though fill = gender has the same function here.
  geom_boxplot(width = 1, outlier.shape = 1) +
  labs(y = "Salary", x = "Gender", title = "Boxplot of ME graduate salaries by gender") + 
  theme_minimal()
```

Create a boxplot of `age`, faceted by `gender`.

```{r age-box}
# boxplots of age data by gender
ggplot(salaries, aes(y = age, fill = gender)) +
  geom_boxplot(width = 1, outlier.shape = 1) +
  labs(y = "Age", x = NULL, title = "Boxplot of ME graduate age by gender") + 
  theme_minimal()
```

\newpage

## Question 5: Scatterplot and correlation

Create a scatterplot of `age` (x-axis) and `salary`, differentiating by 
`gender`. 

```{r multivar-scatter}
# scatterplot of salary across age by gender
ggplot(salaries, aes(x = age, y = salary, color = gender)) +
  geom_point(alpha = 0.5) +
  labs(x = "Age", y = "Salary", title = "Salary vs Age of ME grads by Gender") +
  theme_minimal()
```

*Bonus point*: Is there a correlation between an engineer's salary and age?
What is the estimated Pearson correlation coefficient *r*? Run a formal test.

```{r corr-test}
# correlation test

```

\newpage

## Question 6: Cumulative distribution function

Plot the cumulative distribution function of `salary` by `gender`. Adjust the
x-axis with `scale_x_log10(limits = c(5e4, 5e5))` to zoom in a bit. What do you
notice about the salaries for men and women? Hint: Remember there are greater
differences the farther up you go on a log scale axis.

```{r salary-ecdf}
# plot cdf of salary by gender
ggplot(salaries, aes(x = salary, color = gender)) +
  stat_ecdf(geom = "step") +
  scale_x_log10(limits = c(5e4, 5e5)) +
  labs(x = "Salary", y = "Cumulative Fraction", title = "Cumulative Distribution of Salary of ME Grads by gender") +
  theme_minimal()

```

## Question 7: Quantiles

Calculate the quantiles of `salary` by `gender`. You can either subset the data
with `dplyr::filter()` and dataframe assignment, or you can group by, summarize
by quantile, and ungroup. 

*Bonus point*: Assign the output to a dataframe, and use inline code to call
individual values when answering the following questions. Do not let R use
scientific notation in the text output; check the knitted document.

```{r salary-quantiles, message=FALSE}
# calculate quantiles of salary by gender
salaries_tibble <- as_tibble(salaries)
male_salaries <- filter(salaries_tibble, gender == "M")
male_quantiles <- quantile(male_salaries$salary, seq(0, 1, 0.25)) %>%
  print()

female_salaries <- filter(salaries_tibble, gender == "F")
female_quantiles <- quantile(female_salaries$salary, seq(0, 1, 0.25)) %>%
  print()
```

What is the difference in salary between men and women at the median? 

  * Median salary for women is $90,000
  * Median salary for men is $97,000
  * The difference at the median is $7,000
  
At the top percentile (maximum)? 

  * Maximum salary for women is $350,000
  * Maximum salary for men is $1,027,653
  * The difference at the maximum is $677,653 

Do you think there is a salary difference by gender across the pay scale? What
other information would you need to test your hypothesis?

Yes there is a salary difference by gender across the pay scale. To further test my hypothesis it would be helpful to have an equal amount of salary data on both female and male engineers.

## Question 8: Hypothetical analysis

Think about what other variables you would like to include in an hypothetical
analysis. From your perspective, what are the most important individual,
family, and workforce factors related to salary---beyond gender and age?

Other factors that are important are: time at company, time as an engineer, whether they have any advanced degrees, and which engineers have certificates.

\newpage

## Question 9: Recreate plot

Recreate this plot with the `mpg` dataset. Remember to use `?mpg` for 
information on the dataset and the variables. How would you describe the 
correlation between the independent variable and dependent variable? Do you
see any patterns when considering the third variable?

(View R Markdown PDF for image)

```{r mpg-plot}
# call mpg pdf - you need to recreate it 
#knitr::include_graphics("./mpg-ch7-plot.pdf")
mpg <- as.data.frame(mpg)
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point(size = 3, alpha = 0.5) +
  labs(x = "Engine Displacement (liters)", y = "Highway Miles per Gallon", title = "Mileage by Engine Displacement", subtitle = "Data from 1999 and 2008", color = "Car Class") +
  theme_minimal()
  
```

\newpage

# Appendix

```{r show_code, ref.label = knitr::all_labels(), echo = TRUE, eval = FALSE}
```
